---
title: "Platelet Dosing Models"
subtitle: "BMIN503/EPID600 Final Project"
author: "Narek Israelyan"
format: html
editor: visual
number-sections: true
embed-resources: true
execute:
  eval: false
---

------------------------------------------------------------------------

## Overview {#sec-overview}

This project analyzes pediatric platelet transfusion data to determine if a new dosing model can better predict post-transfusion platelet counts. The primary goal is to compare the efficacy of the current weight-based dosing strategy against a new, combined model that incorporates both the patient's weight and the absolute platelet count of the transfused product. I am working with Dr. David Friedman (Clinical Informatics, Transfusion Medicine) and Dr. Stella Chou (Hematology, Transfusion Medicine).

## Introduction {#sec-introduction}

In transfusion medicine, particularly for pediatric patients who have much smaller total blood volumes, accurately dosing of blood products is important for achieving desired therapeutic outcomes. The current standard of care in pediatrics is to dose platelets based on the patient's weight, calculated as a specific volume-to-weight concentration (eg, 5-10 mL/kg). However, this method does not take into account the actual platelet count of the product, which can vary widely (between 2.5 - 4.5 x 10\^11 platelets/bag). This is a piece of a greater problem in transfusion medicine, where there is significant variability between blood products due to varying storage conditions and differences in donor characteristics that lead, ultimately, to two blood products never providing the same therapeutic dose to the patient. This project seeks to address this gap by developing a more robust model for pediatric platelet dosing.

This challenge requires input from clinical informatics, transfusion medicine, pediatrics and hematology. Transfusion medicine will provide the laboratory context and blood products standards, clinical informatics will help to build the appropriate queries and data analysis, while pediatrics and hematology will provide the clinical context and help to narrow the scope appropriately. Discussions with Dr. David Friedman (clinical informatics and transfusion medicine) have helped to narrow the language of the query to extract the appropriate data and exclude unwanted outliers, while discussions with Dr. Stella Chou (hematology and pediatrics) have helped to determine the most commonly used dosing strategies and indications for transfusion, and which patient populations to focus on.

## Methods {#sec-methods}

I utilized SQL to extract the relevant data from CHOP's clinical data warehouse (Helix). See below for steps.

#### **Created a CTE to limit the platelet transfusions to a narrow scope for piloting.**

```{WITH transfusions_in_scope AS (}
    SELECT *
    FROM prod.semantic.blood_bank_transfusion_all
    WHERE product_id = 'PLT'
        AND order_datetime BETWEEN '2023-07-01' AND '2023-12-31'),
```

#### **Created CTEs to find the most recent pre-transfusion and post-transfusion platelet counts using row number functionality, limiting results to a 7 day window before and after transfusions.**

```{pre_CBC_final AS (}
    SELECT
        blood_bank_transfusion_key,
        pre_specimen_taken_date,
        pre_result_date,
        pre_platelet
    FROM (
        SELECT
            t.blood_bank_transfusion_key,
            c.specimen_taken_date AS pre_specimen_taken_date,
            c.result_date AS pre_result_date,
            c.cbc_platelet_count_numeric AS pre_platelet,
            ROW_NUMBER() OVER (
                PARTITION BY t.blood_bank_transfusion_key
                ORDER BY c.specimen_taken_date DESC
            ) AS rn
        FROM transfusions_in_scope t
        INNER JOIN prod.semantic.procedure_order_result_cbc c
            ON t.patient_key = c.patient_key
        WHERE t.product_id = 'PLT'
            AND c.result_date < t.order_datetime
            AND c.specimen_taken_date >= t.order_datetime - INTERVAL '168 HOURS'
            AND c.cbc_platelet_count_numeric IS NOT NULL
    ) AS pre_candidates
    WHERE rn = 1
),
post_CBC_final AS (
    SELECT
        blood_bank_transfusion_key,
        post_specimen_taken_date,
        post_result_date,
        post_platelet
    FROM (
        SELECT
            t.blood_bank_transfusion_key,
            c.specimen_taken_date AS post_specimen_taken_date,
            c.result_date AS post_result_date,
            c.cbc_platelet_count_numeric AS post_platelet,
            ROW_NUMBER() OVER(
                PARTITION BY t.blood_bank_transfusion_key
                ORDER BY c.specimen_taken_date ASC, c.result_date ASC
            ) AS rn
        FROM transfusions_in_scope t
        INNER JOIN prod.semantic.procedure_order_result_cbc c
            ON t.patient_key = c.patient_key
        WHERE
            t.product_id = 'PLT'
            AND c.specimen_taken_date > t.transfusion_start_datetime
            AND c.specimen_taken_date <= t.order_datetime + INTERVAL '168 HOURS'
            AND c.cbc_platelet_count_numeric IS NOT NULL
    ) AS post_candidates
    WHERE rn = 1
),
```

#### **Created CTE to obtain latest patient weight, which is to be used in dosing calculations.**

```{latest_weight AS (}
    SELECT
        patient_key,
        medication_order_create_date,
        dosage_weight_kg
    FROM (
        SELECT
            patient_key,
            medication_order_create_date,
            dosage_weight_kg,
            ROW_NUMBER() OVER(
                PARTITION BY patient_key, CAST(medication_order_create_date AS DATE)
                ORDER BY medication_order_create_date DESC
            ) AS rn
        FROM prod.core.fact_medication_order
        WHERE dosage_weight_kg IS NOT NULL
    ) AS weight_candidates
    WHERE rn = 1
),
```

#### **Created CTE to pull in contextual information regarding the transfusion and ensure that blood units are not associated with multiple encounters.**

```{admin_details AS (}
    SELECT
        b.blood_unit_number,
        b.blood_admin_start_date,
        b.department_name,
        i.primary_dx,
        i.primary_dx_snomed,
        ROW_NUMBER() OVER(
            PARTITION BY b.blood_unit_number, b.blood_admin_start_date
            ORDER BY i.encounter_key DESC
        ) as rn
    FROM prod.semantic.blood_product_administration b
    INNER JOIN prod.semantic.encounter_inpatient i 
        ON b.encounter_key = i.encounter_key
    WHERE EXISTS (
        SELECT 1 FROM transfusions_in_scope t
        WHERE t.unit_no = b.blood_unit_number
          AND t.transfusion_start_datetime = b.blood_admin_start_date
    )
)
```

#### **Final select and join statements.**

```{SELECT}
    t.product_name AS "Product Name",
    t.blood_bank_transfusion_key AS "Transfusion Key",
    t.unit_no AS "Product DIN",
    t.ecode AS "Ecode",
    ad.department_name AS "Department Transfused",
    ad.primary_dx AS "Primary Dx",
    ad.primary_dx_snomed AS "Primary Dx SNOMED",
    pre.pre_specimen_taken_date AS "DateTime: Pre-Transfusion CBC Taken",
    pre.pre_result_date AS "DateTime: Pre-Transfusion CBC Resulted",
    pre.pre_platelet AS "Pre-Transfusion Platelet Count",
    t.order_datetime AS "DateTime: PLT Ordered",
    t.transfusion_start_datetime AS "DateTime: PLT Transfused",
    t.available_quantity AS "Transfused PLT Volume",
    w.dosage_weight_kg AS "Weight (kg)",
    post.post_specimen_taken_date AS "DateTime: Post-Transfusion CBC Taken",
    post.post_result_date AS "DateTime: Post-Transfusion CBC Resulted",
    post.post_platelet AS "Post-Transfusion Platelet Count"
FROM prod.semantic.blood_bank_transfusion_all t
INNER JOIN admin_details ad
    ON t.unit_no = ad.blood_unit_number
    AND t.transfusion_start_datetime = ad.blood_admin_start_date
    AND ad.rn = 1
LEFT JOIN pre_CBC_final pre
    ON t.blood_bank_transfusion_key = pre.blood_bank_transfusion_key
LEFT JOIN post_CBC_final post
    ON t.blood_bank_transfusion_key = post.blood_bank_transfusion_key
LEFT JOIN latest_weight w
    ON t.patient_key = w.patient_key
    AND CAST(t.order_datetime AS DATE) = CAST(w.medication_order_create_date AS DATE)
ORDER BY
    t.order_datetime;
```

#### **Next, we needed a technologist to manually input the absolute platelet counts as they were sent to us hand-written on a piece of paper that came with the products. After these were completed, the tables were joined to create a final table.**

```{SELECT * FROM PLT.`platelet dosing project2`}
LEFT JOIN PLT.`absolute platelet count`
ON PLT.`platelet dosing project2`.`Product DIN` = PLT.`absolute platelet count`.UnitDin
AND PLT.`platelet dosing project2`.ECODE = PLT.`absolute platelet count`.Ecode
;
```

#### Lastly, using this pilot data, I have began to manipulate and test in R to create regression models and determine the efficacy of models. See below for steps.

```{r}
library(tidyverse)
platelet_data <- read_csv("Platelet Dosing Project NXI.csv")
```

#### Rename columns, ensure correct data type.

```{r}
platelet_data <- platelet_data %>%
  rename(
    Pre_Plt_Count = `Pre-Transfusion Platelet Count`,
    Post_Plt_Count = `Post-Transfusion Platelet Count`,
    Weight_kg = `Weight (kg)`,
    Volume_ml = `Transfused PLT Volume`
  )

platelet_data <- platelet_data %>%
  mutate(
    Pre_Plt_Count = as.numeric(Pre_Plt_Count),
    Post_Plt_Count = as.numeric(Post_Plt_Count),
    Weight_kg = as.numeric(Weight_kg),
    Volume_ml = as.numeric(Volume_ml),
    AbsolutePLT = as.numeric(AbsolutePLT)
  )
```

#### Calculate the key variables (dose and delta).

```{r}
platelet_data <- platelet_data %>%
  mutate(
    Dose_ml_kg = Volume_ml / Weight_kg,
    Delta_Plt = Post_Plt_Count - Pre_Plt_Count
  )
```

#### Filter out rows with missing data.

```{r}
model_data <- platelet_data %>%
  filter(
    !is.na(Post_Plt_Count),
    !is.na(Pre_Plt_Count),
    !is.na(Dose_ml_kg),
    !is.na(AbsolutePLT)
  )
```

#### Build regression models.

```{r}
# Model 1: Current Algorithm (Predicting platelet delta with only dosing)
model_1 <- lm(Delta_Plt ~ Dose_ml_kg, data = model_data)

# Model 2: Unused Algorithm (Predicting platelet delta with only absolute platelet count)
model_2 <- lm(Delta_Plt ~ AbsolutePLT, data = model_data)

# Model 3: Proposed Algorithm (Predicting platelet delta with both models)
model_3 <- lm(Delta_Plt ~ Dose_ml_kg + AbsolutePLT, data = model_data)
```

## Results {#sec-results}

TBD

## Conclusion

TBD
